---
title: "Figure2"
output: html_document
date: "2025-11-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = "data/Fig2/",echo = TRUE)

# Clean workspace
rm(list=ls()) 

set.seed(123)
```

## Figure 2. Oxygen-related pathway heatmaps in digestive tissues  
This figure shows three pathway modules (Antioxidant, Response, Generation) across digestive tissues, visualized as ComplexHeatmap panels with row-level fusion-score barplots.

```{r load_data, echo=TRUE}
matrix_data <- readRDS("../Fig1/matrix_data.rds")
a1          <- read.csv("../Fig1/three_DST_fusion_scores_with_group.csv", row.names = 1)
```



## 1. Build tissue × pathway matrix for oxygen-related signals

```{r build_matrix, echo=TRUE}
library(reshape2)
library(dplyr)

matrix_data_melt <- melt(matrix_data)

all_oxygen_signal_tissue_cluster <- matrix_data_melt %>%
  group_by(tissue, variable) %>%
  summarise(sum_avg_exp_scaled = mean(value), .groups = "drop") %>%
  arrange(desc(sum_avg_exp_scaled))

all_oxygen_signal_tissue_cluster <- dcast(
  all_oxygen_signal_tissue_cluster,
  tissue ~ variable
)

rownames(all_oxygen_signal_tissue_cluster) <- all_oxygen_signal_tissue_cluster[, 1]
all_oxygen_signal_tissue_cluster <- all_oxygen_signal_tissue_cluster[, -1]

# Keep only 1–48 oxygen-related pathways
all_oxygen_signal_tissue_cluster <- all_oxygen_signal_tissue_cluster[, 1:48]

all_oxygen_signal_tissue_cluster_annotation <- a1

digestive <- all_oxygen_signal_tissue_cluster_annotation$tissue %in% c(
  "Rumen","Reticulum","Omasum","Abomasum",
  "Duodenum","Jejum","Ileum","Cecum","Colon","Rectum"
)
```


## 2. Split matrix into three pathway modules

```{r split_matrices, echo=TRUE}
use_antioxidant <- all_oxygen_signal_tissue_cluster[digestive, 1:19]
use_response    <- all_oxygen_signal_tissue_cluster[digestive, 20:32]
use_generation  <- all_oxygen_signal_tissue_cluster[digestive, 33:48]

use_annotation <- all_oxygen_signal_tissue_cluster_annotation[digestive, ]
```


## 3. Create row annotations for tissues

```{r annotations, echo=TRUE}
library(ComplexHeatmap)
library(RColorBrewer)

unique_tissues <- unique(use_annotation$tissue)

tissue_colors <- setNames(
  colorRampPalette(brewer.pal(9, "Set1"))(length(unique_tissues)),
  unique_tissues
)

tissue_annotation <- rowAnnotation(
  tissue = anno_simple(use_annotation$tissue, col = tissue_colors)
)

tissue_factor <- factor(use_annotation$tissue)

col_fun <- colorRampPalette(c("navy", "white", "firebrick3"))(50)
```


## 4. Generate three heatmaps (Antioxidant / Response / Generation)

```{r heatmaps, echo=TRUE}
ht_antioxidant <- Heatmap(
  use_antioxidant,
  name              = "Antioxidant",
  col               = col_fun,
  left_annotation   = tissue_annotation,
  split             = tissue_factor,
  cluster_rows      = TRUE,
  show_row_names    = TRUE,
  show_column_names = TRUE
)

ht_response <- Heatmap(
  use_response,
  name              = "Response",
  col               = col_fun,
  split             = tissue_factor,
  cluster_rows      = TRUE,
  show_row_names    = FALSE,
  show_column_names = TRUE
)

ht_generation <- Heatmap(
  use_generation,
  name              = "Generation",
  col               = col_fun,
  split             = tissue_factor,
  cluster_rows      = TRUE,
  show_row_names    = FALSE,
  show_column_names = TRUE
)
```


## 5. Row barplots for fusion scores

```{r barplots, echo=TRUE}
p2 <- rowAnnotation(Antioxidant = anno_barplot(use_annotation$Antioxidant_fusion))
p3 <- rowAnnotation(Response    = anno_barplot(use_annotation$Response_fusion))
p4 <- rowAnnotation(Generation  = anno_barplot(use_annotation$Generation_fusion))
```


## 6. Draw and save combined figure (PDF, high-resolution landscape)

```{r save_pdf, echo=TRUE,include=TRUE}
pdf("Figure2_complex_heatmap.pdf", width = 14, height = 8)

draw(
  p2 + p3 + p4 + ht_antioxidant + ht_response + ht_generation,
  heatmap_legend_side    = "right",
  annotation_legend_side = "right"
)

dev.off()
p2 + p3 + p4 + ht_antioxidant + ht_response + ht_generation
```


## 7. Export processed matrices and annotations

```{r export_csv, echo=TRUE}
write.csv(all_oxygen_signal_tissue_cluster,
          "oxygen_pathways_mean_by_tissue.csv", row.names = TRUE)

write.csv(all_oxygen_signal_tissue_cluster_annotation,
          "oxygen_pathways_tissue_annotation.csv", row.names = TRUE)

write.csv(use_antioxidant,
          "GI_heatmap_matrix_antioxidant.csv", row.names = TRUE)

write.csv(use_generation,
          "GI_heatmap_matrix_generation.csv", row.names = TRUE)

write.csv(use_response,
          "GI_heatmap_matrix_response.csv", row.names = TRUE)


write.csv(use_annotation,
          "GI_heatmap_annotation.csv", row.names = TRUE)
```


