---
title: "Figure3"
output: html_document
date: "2025-11-24"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = "data/Fig3/",echo = TRUE)

# Clean workspace
rm(list=ls()) 

set.seed(123)
```

# Figure S13

## 1. Data preparation

```{r data_prepare, message=FALSE, warning=FALSE}
# Working directory


# Load marker genes
marker <- read.csv(
  "high_oxidative_cells_kidney_heart_liver_forestomach_marker.csv"
)

# Filter markers
marker <- marker[marker$avg_log2FC > 1, ]


```

## 2. GO enrichment analysis for each tissue / cluster

```{r go_enrichment, message=FALSE, warning=FALSE}
# Load packages
library(clusterProfiler)
library(org.Bt.eg.db)
go_enrich <- list()
go_gene   <- list()

for (i in unique(marker$cluster)) {
  gene_i <- marker[marker$cluster == i, "gene"]
  
  ego_i <- enrichGO(
    gene          = gene_i,
    OrgDb         = org.Bt.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff  = 0.05,
    qvalueCutoff  = 0.2,
    readable      = TRUE
  )
  
  go_enrich[[i]] <- ego_i
  go_gene[[i]]   <- gene_i
}
```

## 3. Dotplot visualization

```{r go_dotplot, fig.width=7, fig.height=5}

if (!is.null(go_enrich$Heart)) {
  dotplot(go_enrich$Heart, showCategory = 20) +
    ggplot2::ggtitle("GO BP enrichment - Heart")
}

if (!is.null(go_enrich$Kidney)) {
  dotplot(go_enrich$Kidney, showCategory = 20) +
    ggplot2::ggtitle("GO BP enrichment - Kidney")
}

if (!is.null(go_enrich$Forestomach_epithelial)) {
  dotplot(go_enrich$Forestomach_epithelial, showCategory = 20) +
    ggplot2::ggtitle("GO BP enrichment - Forestomach epithelium")
}

if (!is.null(go_enrich$liver)) {
  dotplot(go_enrich$liver, showCategory = 20) +
    ggplot2::ggtitle("GO BP enrichment - Liver")
}
```

## 4. Network plot for forestomach epithelium

```{r go_network, message=FALSE, warning=FALSE, fig.width=7, fig.height=6}
if (!is.null(go_enrich$Forestomach_epithelial)) {
  goplot(go_enrich$Forestomach_epithelial, showCategory = 20)
}
```

## 5. Compare GO enrichment across tissues

```{r compare_cluster_go, message=FALSE, warning=FALSE}
xx1 <- compareCluster(
  geneCluster    = go_gene,
  fun            = "enrichGO",
  OrgDb          = org.Bt.eg.db,
  keyType        = "SYMBOL",
  ont            = "BP",
  pAdjustMethod  = "BH",
  pvalueCutoff   = 0.05,
  qvalueCutoff   = 0.2,
  readable       = TRUE
)

xx1 <- pairwise_termsim(xx1)
```

## 6. Enrichment map (emapplot)

```{r emapplot_go, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
p5 <- emapplot(
  xx1,
  pie          = "count",
  layout       = "kk",
  showCategory = 10,
  cex_line     = 0.1
)

p5
```

## 7. Save results

```{r save_results, eval=FALSE}

# Save GO enrichment results
saveRDS(
  go_enrich,
  "GO_results_4_tissues.rds"
)


# Export compareCluster results
write.csv(
  xx1@compareClusterResult,
  "Cluster_compare_network.csv",
  row.names = FALSE
)
```




## Figure 3A

## Figure 3A. Antioxidant activity across tissues
## Figure 3A. Antioxidant-related pathways across tissues

```{r Figure3A_violin_three_paths, echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
library(dplyr)
library(FSA)
library(ggplot2)
library(tidyr)

# Load metadata
metadata <- read.csv(
  "Figure3A_high_oxidative_cells_kidney_heart_liver_forestomach_metadata.csv",
  header = TRUE,
  row.names = 1
)

# Extract the three pathways
df <- metadata[, c(
  "tissue",
  "tissue_cluster",
  "Antioxidant_activity1",
  "peroxiredoxin.activity1",
  "peroxidase.activity1"
)]

df$tissue <- factor(df$tissue)

# Reshape into long format for multi-panel violin plot
df_long <- df %>%
  pivot_longer(
    cols = c(
      Antioxidant_activity1,
      peroxiredoxin.activity1,
      peroxidase.activity1
    ),
    names_to  = "pathway",
    values_to = "score"
  )

# Rename pathway labels for prettier plotting
df_long$pathway <- factor(
  df_long$pathway,
  labels = c(
    "Antioxidant activity",
    "Peroxiredoxin activity",
    "Peroxidase activity"
  )
)

# Violin + boxplot + jitter
p_violin_three <- ggplot(
  df_long,
  aes(x = tissue, y = score, fill = tissue)
) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.8) +
  facet_wrap(~ pathway, scales = "free_y", nrow = 1) +
  labs(
    x = "Tissue",
    y = "Activity score",
    title = "Figure 3A. Antioxidant-associated pathways across tissues"
  ) +
  theme_minimal() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    strip.text      = element_text(size = 12, face = "bold")
  )

p_violin_three
```





## Figure 3B. Energy metabolism pathway activities across tissues

```{r Figure3B_setup, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(reshape2)
library(ggplot2)
library(pheatmap)
library(FSA)

# Load precomputed pathway scores
tissue_score <- read.csv(
  "Figure3B_high_oxidative_cells_kidney_heart_liver_forestomach_energy_metabolism_scores.csv",
  row.names = 1, header = TRUE
)

# Select antioxidant + energy metabolism pathways (columns 64–87 may vary)
tissue_score1 <- tissue_score[, c("Antioxidant_activity1",colnames(tissue_score)[64:87])]
```


## 1. Heatmap of mean pathway activity across tissues

```{r Figure3B_heatmap, echo=TRUE, message=FALSE, warning=FALSE}
# Compute mean activity per tissue
tissue_mean <- tissue_score1 %>%
  group_by(tissue) %>%
  summarise(across(everything(), mean, na.rm = TRUE))

tissue_mean <- as.data.frame(tissue_mean)
rownames(tissue_mean) <- tissue_mean[, 1]
tissue_mean <- tissue_mean[, -1]

# Heatmap
pheatmap::pheatmap(tissue_mean)
```

---

## 2. Circular barplot with error bars (polar coordinate plot)

```{r Figure3B_circular_plot, echo=TRUE, message=FALSE, warning=FALSE}
# Reshape data
tissue_score2 <- melt(tissue_score1)
colnames(tissue_score2)[2] <- "pathways"

# Compute mean and standard error by tissue × pathway
tissue_stats <- tissue_score2 %>%
  group_by(tissue, pathways) %>%
  summarise(
    value_mean = mean(value, na.rm = TRUE),
    value_se   = sd(value, na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  arrange(pathways) %>%
  mutate(midpoint = value_mean + 1)

# Circular barplot with SE
ggplot(tissue_stats, aes(x = factor(pathways), y = value_mean + 1)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(
    aes(ymin = value_mean - value_se + 1,
        ymax = value_mean + value_se + 1),
    width = 0.2
  ) +
  geom_line(aes(y = midpoint, group = 1), color = "black", size = 1) +
  facet_wrap(~ tissue, scales = "free_y") +
  coord_polar(start = 0) +
  labs(
    title = "Energy metabolism pathway activities",
    x = "Pathway",
    y = "Activity (mean ± SE)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6)
  )
```

---

## 3. Dunn’s post-hoc test for one selected pathway

```{r Figure3B_dunn_test, echo=TRUE, message=FALSE, warning=FALSE}
## Perform Dunn test for all pathways in tissue_score1
library(FSA)

# Pathway columns = all except "tissue"
pathway_cols <- setdiff(colnames(tissue_score1), "tissue")

# Create output directory (optional)
dir.create("Figure3B_Dunn_results", showWarnings = FALSE)

# Loop through each pathway
for (p in pathway_cols) {
  
  message("Processing pathway: ", p)
  
  # Build formula dynamically
  formula_p <- as.formula(paste0(p, " ~ tissue"))
  
  # Run Dunn test
  dt <- dunnTest(
    formula_p,
    data   = tissue_score1,
    method = "bonferroni"
  )
  
  # Extract results table
  res <- dt$res
  
  # Build a clean filename
  clean_name <- gsub("[^A-Za-z0-9_.-]", "_", p)
  output_file <- file.path(
    "Figure3B_Dunn_results",
    paste0(clean_name, "_Dunn_bonferroni.csv")
  )
  
  # Save CSV
  write.csv(res, output_file, row.names = FALSE)
}

```







